### 2019-10-21 Manuscript/Manuscript.tex
Updates from Overleaf

\begin{flushleft}
{\Large \textbf{Emergence of Indonesia and New Guinea as a driver for Neogene cooling}}

@@

Indonesia and New Guinea (I\&NG) have an outsized contribution to modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic outgassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a}. Here we use geologic data to reconstruct the shorelines of I\&NG over the past 15~m.y. (Fig. \ref{fig:shoreline_growth}) and quantify the increase in land area. We use these paleoshorelines as boundary conditions within a coupled weathering-climate model to quantify the change in steady-state \pCOtwo associated with this emergence (Fig. \ref{fig:scenario_pCO2}). We find that the modeled decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in I\&NG and enhanced global weatherability, we use the spatially-resolved GEOCLIM model \cite{Godderis2014a, Godderis2017c}. GEOCLIM consists of a silicate weathering model linked to climate model runs at different \pCOtwo levels, allowing for estimates of long-term steady-state \pCOtwo associated with coupled changes in weathering sinks and climatology.

@@

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with the emergence of I\&NG. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006a, Delworth2006b} (see SI). The primary strength of these experiments for this analysis is that all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of \pCOtwo to be isolated. We use terrestrial and marine sedimentary deposits to determine the position of paleoshorelines of I\&NG over the past 15~m.y. and include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (Fig. \ref{fig:shoreline_growth}; see SI). In New Guinea, emergence in the Middle Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{vanUfford2005a, Cloos2005a, Baldwin2012a}, which drove exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. Exhumation of the New Guinea Central Range accelerated over the past 4~m.y. associated with slab-breakoff and buoyant uplift \cite{Cloos2005a} while in eastern New Guinea, progressive jamming of the north-dipping subduction zone also caused major uplift over this time interval \cite{vanUfford2005a}. These tectonic drivers and other throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of islands in the area at ca. 15, 10, and 5~Ma (``reduced I\&NG + Sunda Shelf'' scenarios; Fig. \ref{fig:scenario_pCO2}). Given that the Sunda Shelf is presently submerged, across the shelf we distribute felsic and siliciclastic lithologies and assign shallow slope values in a manner that is consistent with existing observations, we also extrapolate the runoff field from surrounding lands (see SI). While we modify the land area, the position of the tectonic blocks remain fixed since we use a climate model forced with modern geography. While there has been motion of these tectonic blocks since 15~Ma, they have remianed within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (see SI). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed I\&NG'' scenario; Fig. \ref{fig:scenario_pCO2}).

Using the 3,339 unique parameter combinations, the ``reduced I\&NG + Sunda Shelf'' scenarios resulted in 533--611~ppm, 466--502~ppm, and 394--412~ppm \pCOtwo for 15, 10, and 5~Ma respectively (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of I\&NG, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$533--611~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a, Mejia2017a, Ji2018a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (see SI), indicating that the increase in subaerially-exposed land area and tectonic topography of I\&NG is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica followed by Northern Hemisphere ice sheets developing in the Pliocene as \pCOtwo subsequently decreased.

The results of our I\&NG scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of lower-slope basaltic lavas on steady-state \pCOtwo \cite{Kent2008a, Kent2013a}, we replace mafic volcanics associated with the Deccan and Ethiopian Traps with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling\cite{Kent2008a, Kent2013a}, its contribution is less than that of the wetter and higher-relief I\&NG.

Other hypotheses to explain ice sheet growth over the Neogene invoke changes in ocean/atmosphere circulation including: further climatic isolation of Antarctica due to strengthening of the circumpolar current \cite{Shevenell2004a}; increased atmospheric moisture in the Northern Hemisphere due to intensified thermohaline circulation following Panama Isthmus emergence \cite{Haug1998a}; and cooling of North America resulting from a strengthened Walker Circulation associated with I\&NG emergence \cite{Molnar2015a}. Such changes in ocean/atmosphere circulation are likely to modulate \pCOtwo thresholds for glacial initiation and ice sheet growth \cite{DeConto2008a}. However, the prolonged timescale of the cooling trend since 15~Ma (Fig. \ref{fig:shoreline_growth}C) is most readily attributable to decreasing \pCOtwo associated with evolving geological sources and sinks of carbon, modulated by the silicate weathering feedback \cite{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}.

Uplift in the Himalayas \cite{Raymo1988a} or a decrease in volcanic outgassing \cite{Berner1983a} have also been proposed as drivers for Neogene cooling, although recent analyses show Cenozoic seafloor spreading to be relative uniform \cite{Rowan2016a}. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalaya lithologies \cite{Myrow2015a, Colleps2018a}, but could also be partially driven by the exhumation of mafic and organic-rich forearcs sediments in I\&NG during arc-continent collision.

In addition to I\&NG, paleogeographic changes elsewhere on Earth, particularly in the Himalaya and Central America, would have effected geological carbon sources and sinks. Yet, not only does the history of I\&NG emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but also our coupled weathering-climate model indicates that the associated \pCOtwo change is sufficient to explain the magnitude of cooling. These results highlight that the Earth's climate state is particularly sensitive to tropical paleogeography.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{Figures/shoreline_growth.pdf}
    \caption{The emergence of Indonesia and New Guinea (I\&NG) from the mid-Miocene to present in the region referred to in climate science literature as the Maritime Continent. Past shorelines are shown in A with associated area estimates summarized in B (where I\&NG refers to the sum of all the subdivided regions). A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in C.}
    \label{fig:shoreline_growth}
\end{figure}

>>>>

\begin{flushleft}
{\Large \textbf{Emergence of the SE Asian islands as a driver for Neogene cooling}}

@@

The SE Asian islands (SEAI) have an outsized contribution on modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic outgassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a}. Here we use geologic data to reconstruct the shorelines of SEAI over the past 15~m.y. (Fig. \ref{fig:shoreline_growth}) and quantify the increase in land area. We use these paleoshorelines as boundary conditions within a coupled weathering-climate model to quantify the change in steady-state \pCOtwo associated with this emergence (Fig. \ref{fig:scenario_pCO2}). We find that the modeled decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in SEAI and enhanced global weatherability, we use the spatially-resolved GEOCLIM model \cite{Godderis2014a, Godderis2017c}. GEOCLIM consists of a silicate weathering model linked to climate model runs at different \pCOtwo levels, allowing for estimates of long-term steady-state \pCOtwo associated with coupled changes in weathering sinks and climatology.

@@

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with the emergence of SEAI. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006a, Delworth2006b} (see SI). The primary strength of these experiments for this analysis is that all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of \pCOtwo to be isolated. We use terrestrial and marine sedimentary deposits to determine the position of paleoshorelines of SEAI over the past 15~m.y. and include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (Fig. \ref{fig:shoreline_growth}; see SI). In New Guinea, emergence in the Middle Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{vanUfford2005a, Cloos2005a, Baldwin2012a}, which drove exhumation of the 1440~km long Irian-Marum-April Ophiolite Belt. Exhumation of the New Guinea Central Range accelerated over the past 4~m.y. associated with slab-breakoff and buoyant uplift \cite{Cloos2005a} while in eastern New Guinea, progressive jamming of the north-dipping subduction zone also caused major uplift over this time interval \cite{vanUfford2005a}. These tectonic drivers and other throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of islands in the area at ca. 15, 10, and 5~Ma (``reduced SEAI + Sunda Shelf'' scenarios; Fig. \ref{fig:scenario_pCO2}). Given that the Sunda Shelf is presently submerged, across the shelf we distribute felsic and siliciclastic lithologies and assign shallow slope values in a manner that is consistent with existing observations, we also extrapolate the runoff field from surrounding lands (see SI). While we modify the land area, the position of the tectonic blocks remain fixed since we use a climate model forced with modern geography. While there has been motion of these tectonic blocks since 15~Ma, they have remianed within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (see SI). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

Using the 3,339 unique parameter combinations, the ``reduced SEAI + Sunda Shelf'' scenarios resulted in 533--611~ppm, 466--502~ppm, and 394--412~ppm \pCOtwo for 15, 10, and 5~Ma respectively (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$533--611~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a, Mejia2017a, Ji2018a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (see SI), indicating that the increase in subaerially-exposed land area and tectonic topography of SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica followed by Northern Hemisphere ice sheets developing in the Pliocene as \pCOtwo subsequently decreased.

The results of our SEAI scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of lower-slope basaltic lavas on steady-state \pCOtwo \cite{Kent2008a, Kent2013a}, we replace mafic volcanics associated with the Deccan and Ethiopian Traps with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling\cite{Kent2008a, Kent2013a}, its contribution is less than that of the wetter and higher-relief SEAI.

Other hypotheses to explain ice sheet growth over the Neogene invoke changes in ocean/atmosphere circulation including: further climatic isolation of Antarctica due to strengthening of the circumpolar current \cite{Shevenell2004a}; increased atmospheric moisture in the Northern Hemisphere due to intensified thermohaline circulation following Panama Isthmus emergence \cite{Haug1998a}; and cooling of North America resulting from a strengthened Walker Circulation associated with SEAI emergence \cite{Molnar2015a}. Such changes in ocean/atmosphere circulation are likely to modulate \pCOtwo thresholds for glacial initiation and ice sheet growth \cite{DeConto2008a}. However, the prolonged timescale of the cooling trend since 15~Ma (Fig. \ref{fig:shoreline_growth}C) is most readily attributable to decreasing \pCOtwo associated with evolving geological sources and sinks of carbon, modulated by the silicate weathering feedback \cite{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}.

Uplift in the Himalayas \cite{Raymo1988a} or a decrease in volcanic outgassing \cite{Berner1983a} have also been proposed as drivers for Neogene cooling, although recent analyses show Cenozoic seafloor spreading to be relative uniform \cite{Rowan2016a}. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalaya lithologies \cite{Myrow2015a, Colleps2018a}, but could also be partially driven by the exhumation of mafic and organic-rich forearcs sediments in SEAI during arc-continent collision.

In addition to SEAI, paleogeographic changes elsewhere on Earth, particularly in the Himalaya and Central America, would have effected geological carbon sources and sinks. Yet, not only does the history of SEAI emergence coincide with Neogene cooling and the onset of Northern Hemisphere glaciation, but also our coupled weathering-climate model indicates that the associated \pCOtwo change is sufficient to explain the magnitude of cooling. These results highlight that the Earth's climate state is particularly sensitive to tropical paleogeography.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the SE Asian islands (SEAI) from the mid-Miocene to present in the region referred to in climate science literature as the Maritime Continent. Past shorelines are shown in A with associated area estimates summarized in B. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in C.}
    \label{fig:shoreline_growth}
\end{figure}
